---
title: "Untitled"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plotly)
library(UsingR)
library(usethis)
library(dplyr)
library(tidyr)
library(sampling)
library(sm)
library(stringr)
# Ray's libraries
library(maps)
library(mapproj)
library(mapdata)
library(ggplot2)

?as.integer

### PRE-PROCESSING ###
# load and transform datatypes where required ###
#data <- read.csv('survey_results_responses.csv')
data$TotalComp<-as.integer(data$TotalComp)
data$YearsCode<-as.integer(data$YearsCode)
data$YearsCodePro<-as.integer(data$YearsCodePro)
#Analysis of US Responses only where Total Compensation > US Federal Minimum Wage 2021
us_data<-subset(data,Country == 'United States of America'&
                Currency == 'USD\tUnited States dollar'&
                  CompTotal > 15000 &
                (Gender == 'Man' | Gender == 'Woman'))
#search for and remove  outliers in Salary data using 1.5 IQR
f<-fivenum(us_data$CompTotal);f
subset(us_data, CompTotal > f[4] + 1.5*(f[4] - f[2])) 
us_data<-subset(us_data,CompTotal<f[4]+1.5*(f[4]-f[2])) 

```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
